ECCCos from the Black Box:\\ Faithful Explanations through\\ Energy-Constrained Conformal Counterfactuals

### 2023-05-16 paper/paper.tex

\begin{minipage}[c]{0.50\textwidth}
  \centering
  \includegraphics[width=0.8\textwidth]{../artifacts/results/images/poc.png}
  \captionof{figure}{An example involving linearly separable synthetic data and illustrating how ECCCo compares to Wachter, Schut and REVISE. The factual class is 2 and the target class is 1. Contours indicate the predicted probability given by a Joint Energy Model that the counterfactual belongs to the target class.}\label{fig:poc}
\end{minipage}  
\hfill
\begin{minipage}[c]{0.40\textwidth}
\captionof{algorithm}{Generating ECCCos (For more details, see Appendix~\ref{app:eccco})}\label{alg:eccco}
\begin{algorithmic}[1]
  \Require $\mathbf{x}, \mathbf{y}^+, M_{\theta}, f, \Lambda, \alpha, \mathcal{D}, T, \eta, n_{\mathcal{B}}, N_{\mathcal{B}}$ \linebreak where $M_{\theta}(\mathbf{x})\neq\mathbf{y}^+$
  \Ensure $\mathbf{x}^\prime$
  \State Initialize $\mathbf{z}^\prime \gets f^{-1}(\mathbf{x})$
  \State Generate buffer $\mathcal{B}$ of $N_{\mathcal{B}}$ conditional samples $\hat{\mathbf{x}}_{\theta}|\mathbf{y}^+$ using SGLD (Equation~\ref{eq:sgld})
  \State Run \textit{SCP} for $M_{\theta}$ using $\mathcal{D}$
  \State Initialize $t \gets 0$
  \While{\textit{not converged} or $t < T$}
  \State $\hat{\mathbf{x}}_{\theta, t} \gets \text{rand}(\mathcal{B},n_{\mathcal{B}})$
  \State $\mathbf{z}^\prime \gets \mathbf{z}^\prime - \eta \nabla_{\mathbf{z}^\prime} \mathcal{L}(\mathbf{z}^\prime,\mathbf{y}^+,\hat{\mathbf{x}}_{\theta, t}; \Lambda, \alpha)$
  \State $t \gets t+1$
  \EndWhile
  \State $\mathbf{x}^\prime \gets f(\mathbf{z}^\prime)$
\end{algorithmic}
\end{minipage}

% \begin{minipage}[c]{\textwidth}
%   \centering
%   \includegraphics[width=0.8\textwidth]{../artifacts/results/images/mnist_eccco.png}
%   \captionof{figure}{Original image (left) and ECCCos for turning a 9 (nine) into a 7 (seven) for different black-box models from left to right: Multi-Layer Perceptron (MLP), Ensemble of MLPs, Joint Energy Model (JEM), Ensemble of JEMs.}\label{fig:eccco-mnist}
% \end{minipage}

>>>>

\begin{minipage}[c]{0.45\textwidth}
  \centering
  \includegraphics[width=1.0\textwidth]{../artifacts/results/images/poc_gradient_fields.png}
  \captionof{figure}{Gradient fields and counterfactual paths for different generators. The objective is to generate a counterfual in the `blue' class for a sample from the `orange' class. Bright yellow stars indicate conditional samples generated through SGLD. The underlying classifier is a Joint Energy Model.}\label{fig:poc}
\end{minipage}  
\hfill
\begin{minipage}[c]{0.45\textwidth}
\captionof{algorithm}{Generating ECCCos (For more details, see Appendix~\ref{app:eccco})}\label{alg:eccco}
\begin{algorithmic}[1]
  \Require $\mathbf{x}, \mathbf{y}^+, M_{\theta}, f, \Lambda, \alpha, \mathcal{D}, T, \eta, n_{\mathcal{B}}, N_{\mathcal{B}}$ \linebreak where $M_{\theta}(\mathbf{x})\neq\mathbf{y}^+$
  \Ensure $\mathbf{x}^\prime$
  \State Initialize $\mathbf{z}^\prime \gets f^{-1}(\mathbf{x})$
  \State Generate buffer $\mathcal{B}$ of $N_{\mathcal{B}}$ conditional samples $\hat{\mathbf{x}}_{\theta}|\mathbf{y}^+$ using SGLD (Equation~\ref{eq:sgld})
  \State Run \textit{SCP} for $M_{\theta}$ using $\mathcal{D}$
  \State Initialize $t \gets 0$
  \While{\textit{not converged} or $t < T$}
  \State $\hat{\mathbf{x}}_{\theta, t} \gets \text{rand}(\mathcal{B},n_{\mathcal{B}})$
  \State $\mathbf{z}^\prime \gets \mathbf{z}^\prime - \eta \nabla_{\mathbf{z}^\prime} \mathcal{L}(\mathbf{z}^\prime,\mathbf{y}^+,\hat{\mathbf{x}}_{\theta, t}; \Lambda, \alpha)$
  \State $t \gets t+1$
  \EndWhile
  \State $\mathbf{x}^\prime \gets f(\mathbf{z}^\prime)$
\end{algorithmic}
\end{minipage}
